\documentclass[letterpaper,draftclsnofoot,10pt,onecolumn,titlepage]{IEEEtran}\usepackage[margin=0.75in]{geometry}
\renewcommand\thesection{\arabic{section}}
\renewcommand\thesubsection{\thesection.\arabic{subsection}}
\renewcommand\thesubsubsection{\thesubsection.\arabic{subsubsection}}
\renewcommand\thesectiondis{\arabic{section}}
\renewcommand\thesubsectiondis{\thesectiondis.\arabic{subsection}}
\renewcommand\thesubsubsectiondis{\thesubsectiondis.\arabic{subsubsection}}

\usepackage{graphicx}
\usepackage{url}
\usepackage{float}
\usepackage{color}
\usepackage{graphicx}
\usepackage{enumitem}
\usepackage[justification=centering]{caption}
\usepackage{listings}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}
\lstset{language=C++,
        basicstyle=\ttfamily,
        keywordstyle=\color{blue}\ttfamily,
        stringstyle=\color{red}\ttfamily,
        commentstyle=\color{green}\ttfamily,
        morecomment=[l][\color{magenta}]{\#},
        backgroundcolor=\color{backcolour}
}
\usepackage{geometry}
\graphicspath{{../img/}}
\usepackage{parskip}
\setlength{\parindent}{0pt}
\usepackage{hyperref}
\usepackage{geometry}
\linespread{1}
\setcounter{secnumdepth}{5}
\setcounter{secnumdepth}{5}

\title{Writing Assignment 2}

\author{
	Alec Merdler\\
	\and
	CS 444: Operating Systems II\\
	\and
	Spring 2017\\
}

\date{May 17, 2017}

\begin{document}
\begin{titlepage}
\clearpage\maketitle
\thispagestyle{empty}

\maketitle
\end{titlepage}

\section{Introduction}
Computers are fundamentally useless without the ability to receive different inputs and emit unique output.
Whether the data comes from a hard drive, solid state drive, mouse, keyboard, or virtual reality headset,
the operating system needs to be able to translate and deliver the data to the various programs that need
it. Conversely, programs running on the operating system need a way to provide feedback back out to 
devices. The design and implementation of handling I/O differs between three major operating systems: Windows,
FreeBSD, and Linux. However, there are also many similarities amongst them as well.

\section{Windows}
 
\subsection{I/O Components}
The Windows I/O System consists of many executive components that together manage devices.
The Windows operating system uses a number of executive components that work together to manage input and 
output devices connected to the computer \cite{Windows}.

\begin{description}

    \item[Security] 
    Uniform security and naming across devices in order to protect shareable resources.
    
    \item[Asynchronous]
    High performance asynchronous packet based I/O to allow for the creation and maintenance of scalable 
    applications.
    
    \item[Multiple Platforms]
    Services that allow drivers to be written in a high level language and easily ported between different 
    machine architectures.
    
    \item[Extensibility]
    Layering and extensibility to allow for the addition of drivers that tranparently modify the behavior 
    of other drivers or devices, without requiring any changes to the driver whole behavior or device is 
    modified.
    
    \item[Dynamic Loading]
    Dynamic unloading and loading of device drivers so that they can be loaded on demand and will not 
    consume system resources when not in use.
    
    \item[Plug and Play]
    Support for plug and play, where the system locates and installs drivers for newly detected hardware, 
    assigns them hardware resources they require, and also allows applications to discover and activate 
    device interfaces.
    
    \item[Power Management]
    Support for power management so that the system or individual devices can enter low power states.
    
    \item[File Systems]
    Support for multiple installable file systems, including FAT, the CD-ROM file sustem(CDFS), the 
    Universal Disk Format (UDF) file system, and the Windows file system (NTFS).
    
    \item[Driver Tools]
    Windows Management Instrumentation (WMI) support and diagnostics tools so that drivers can be managed 
    and monitored through WMI scripts and programs.
    
\end{description}

\subsection{I/O Processing}
Windows provides a hardware abstraction layer (HAL) at the lowest level in order to encapsulate the different
processor and interrupt signals and instead expose a uniform set of application programming interfaces (APIs)
to device drivers. The next layer is the actual device drivers. This layer is followed by the I/O system
layer, which includes the various executive components listed in the previous section. Most I/O
operations in Windows are synchronous, which means the application thread will block while the device
performs its data operations. Asynchronous I/O operations are enabled by specifying the FILE-FLAG-OVERLAPPED
flag when calling the CreateFile function. This will allow the application to continue executing while the
device performs the I/O operation \cite{Windows}.

\begin{lstlisting}[language=C++]
HANDLE WINAPI CreateFile(
  _In_     LPCTSTR               lpFileName,
  _In_     DWORD                 dwDesiredAccess,
  _In_     DWORD                 dwShareMode,
  _In_opt_ LPSECURITY_ATTRIBUTES lpSecurityAttributes,
  _In_     DWORD                 dwCreationDisposition,
  _In_     DWORD                 dwFlagsAndAttributes,
  _In_opt_ HANDLE                hTemplateFile
); 
\end{lstlisting}

Above is the implementation of the CreateFile function. This function is used to create or open files or
other I/O devices. The Windows scheduler processes the file descriptors using a C-LOOK algorithm. Once
the list contains requests, the drivers will iterate through the list from the lowest to highest sector
\cite{Windows_CreateFile}.

\section{FreeBSD}

\subsection{I/O Components}
There are multiple different types of file descriptors that are used for difference processes. These 
file descriptors types and uses are as follows:

\begin{itemize}
\item  PIPE - pipe
\item  FIFO - named pipe
\item  SEM - POSIX semaphore
\item  PTS - pseudo-teletype master device
\item  DEV - device not reference by a vnode
\item  MQUEUE - POSIX message queue
\item  KQUEUE - event queue
\item  CRYPTO - cryptographic hardware
\item  SHM - POSIX shared memory
\item  PROCDESC - process
\item  VNODE - file or device
\item  SOCKET - communication endpoint
\end{itemize}

These file descriptors are important because each data structure that is used to implement the 
I/O scheduler has one of these types. Also, there are multiple different flags that go along 
with these file descriptors that will tell the operating system whether to block or not or if 
the process will be asynchronous or synchronous \cite{FreeBSDBook}.

\subsection{I/O Processing}
Below is the implementation of the standard disk scheduler in FreeBSD. A significant difference between 
this scheduler and the Linux scheduler is the ability to delete, load, and unload. Delete is used to free
resources when a scheduler is not in use. Load/unload are used can hook into the respective scheduler events
to perform specific operations. Every time init is called, it will allocate a struct containing device 
information, which is referenced using a file pointer \cite{FreeBSDBook}.

\begin{lstlisting}[language=C++]
struct _disk_sched_interface {
       struct _disk_sched_interface *next;
       char *name;     /* symbolic name */
       int refcount;   /* users of this scheduler */

       void (*disksort)(struct buf_queue_head *head, struct buf *bp);
       void (*remove)(struct buf_queue_head *head, struct buf *bp);
       struct buf *(*get_first)(struct buf_queue_head *head);
       void (*init)(struct buf_queue_head *head);
       void (*delete)(struct buf_queue_head *head);
       void (*load)(void);     /* load scheduler */
       void (*unload)(void);   /* unload scheduler */
};
\end{lstlisting}\cite{FreeBSD}

\section{Conclusion}
The design and implementation of I/O in FreeBSD is very similar to Linux, in that they both use C-LOOK algorithm.
All three operating systems provide file descriptors to specify asynchronous data operations, but only FreeBSD
requires the type of descriptor. Windows has several layers of abstraction, including the hardware 
abstraction layer (HAL) to provide a uniform interface for all device drivers across different hardware 
architectures. This is similar to the Linux kernel itself, which provides a uniform set of APIs across 
countless devices. FreeBSD also has a form of HAL.

\bibliographystyle{IEEEtran}
\bibliography{biblio}

\end{document}
